Measuring contrast processing in the visual system using SSVEP

### 2024-11-28 responseToReviews.md

Adaptation effects.  Adapted contrast response functions have been measured by Nelson et al., (EEG Journal, 1984) by comparing CRFs measured by sweeping contrast from low to high vs high to low. They found that responses at low contrast were smaller in the “down” sweeps and attributed it to differences in prior adaptation.

> we could not find this reference. We did find an excellent (and sometimes humorous) review of SSVEPs by Nelson (Nelson, J. I., Seiple, W. H., Kupersmith, M. J., & Carr, R. E. (1984). Lock-in Techniques for the Swept Stimulus Evoked Potential. Journal of Clinical Neurophysiology, 1(4), 409–436. doi:10.1097/00004691-198401040-00002) but could not find a mention of adaptation effects.

Other studies also measured adaptation effects on the CRF Heinrich and Bach, (IOVS, 2001), Ross and Speed (1991, Proc Roy Soc). And the hard-to-find Bach, Greenlee and Buhler (Clin Vis Sci, 1996) paper that modeled the effect within a contrast gain control framework.

> The Ross and Speed paper measures masking using SSVEP but other adaptation measures are, as far as we can tell, psychophysical.  Heinrich and Bach are concerned with transient changes in SSVEP amplitude within a recording which may have an adaptation origin but which manifest as response increases over the first few seconds and seems a little mysterious (we often observe these as well). Bach, Greenlee and Buhler is, however, pretty unequivocal and we now include it. We like the suggestion that the  'increase' in response is attributable to a shift from a square wave response to a more sinusoidal waveform as the apparent contrast of the stimulus is decreased by adaptation-driven gain control. The response power which was distributed across odd harmonics in the original response can now be concentrated in the 1st harmonic after adaptation.

@@

Gain control development.  Morrone and Burr (Nature, 1986) used frequency tagging to measure orientation dependence of masking in infants and adults, as did Candy et al. The difference in the two studies is that in adults Morrone and Burr found response gain for cross-oriented maskers but Candy et al found contrast gain.   Skoczenski and Norcia used a high contrast 30 Hz dynamic 2D noise masker and variable contrast reversing gratings. They found masking consistent with a contrast gain effect.

> We have clarified that these papers measure different types of gain control

Page 5 bottom. The reciprocal effect of test and masker was first shown by Regan and Regan  (1988; cited)  for flicker modulation depth and then for gratings by Brown et al., (IOVS,  1999 for dichoptic masking) and  then  by candy et al., (J. Neurosci, 2001) along with Busse et al. (Neuron, 2009) as cited.

> Thanks - we have included these references

 It’s good to have cited Regan and Regan paper that introduced the notion of fingerprints in the spectrum being tied to the nature of the non-linearity. Maybe expand their “fingerprint” approach a little more.

@@

>

Figure 4. Oversaturation occurs because the denominator overtakes the numerator at high contrast as it has two terms in it, not the single term in the numerator. Would be worth showing total power not saturating. Is this a general effect?

> We think this topic is probably worth a more detailed treatment than we can provide here. Broadly, much of the oversaturation effect can be modeled by an increase in saturation and therefore harmonic distortion - power is transferred from the first to higher harmonics as contrast increases. It seems likely that total power does not saturate (perceptually, there appears to be a monotonic increase in contrast for example and discrimination thresholds increase up to 100%) but the issue is complicated by the fact that the distribution of neurons sensitive to different contrast levels might not be uniform and that neurons might alter their response properties to signal with optimum sensitivity around a particular range. Our goal here is simply to show people new to the field how the CRF is approximately log-linear with contrast.

@@

Equation 3 is a special case where the neurons in the numerator are so selective that they don’t “see” the masker except by being influenced indirectly by the gain pool.  Putting the test and mask in the numerator and summing them before exponentiating yields robust IM. Weights can be used to reflect the neurons preference for the test vs the mask.  Does equation 3 produce IM?

> In the accompanying code we specifically allow the user to alter the way that normalization is computed and many of the (commented) options produce robust IM. We encourage the user to experiment with different configurations to explore their effects. We have updated the text to make this more explicit.

@@

Page 8 figure 5. The extrapolation to zero amplitude method should be attributed to Cambell and Maffei (J Physiol) 1970). Campbell and Kulikowski (J Physiol, 1972) made a detailed comparison to psychophysics.

> We have now included these references

Figure 5 is an oversimplification. As it doesn’t take into account how the additive noise mixes with the internal VEP response.  This was modeled in Norcia et al., (Vision Res, 1989). The additive noise adds a bias to the measurement, especially if the slope of the response function is shallow. The authors should try to connect the linear regression approach for threshold measurement to the sigmoidal non-linearity, as they are different models of the CRF. In practice the difference at the low contrast end is probably too small to matter and the saturation region is excluded.

> We have now note that it is important to consider noise for situations where the SNR is low or the slope is shallow and we expand on the point that you have to stick to the part of the CRF where the curve is more or less linear - as we showed explicitly in the figure.

@@

Page 9. The sweep method was first developed by David Regan (Vision Res, 1973) for use in objective refraction.

> We have included this reference.

Page 9 “This approach” seems to refer to the sweep VEP, not the method of measuring at multiple contrast levels over a series of trials which was used in the Atkinson/Braddick papers. The sweep method refers to the approach of incrementing or decrementing the stimulus value within a single trial and then averaging the sweeps to get the response function.  Regan argued that this was a better way of estimating response functions because amplitude variations over a session are distributed over the whole response function shifting it up or down instead of distorting its shape (Regan (Vison Res, 1975, color coding).

> This is a good point but the (simpler) one we are trying to make here is that extrapolating from high stimulus levels down to zero provides a robust estimate of visual sensitivity in subjects where such measurements might otherwise be hard to make. We have altered the text to clarify this.

### 2024-11-28 review.qmd
Now on 'thresholds'

To illustrate the effect of contrast gain control [@Heeger1992], we next include a second component (a 'mask') that contributes to the gain pool of the first (the 'target'). The mask stimulus will suppress the response at the target frequency, reducing its amplitude. The suppression is reciprocal - activity at the mask frequency is also reduced by the presence of the target, in a contrast-dependent manner [@Busse2009]. At a single pair of (matched) contrast levels, we see a complex pattern of intermodulation terms in the transduced waveforms (left panel of @fig-twoinputs) and in the Fourier spectrum (middle panel of @fig-twoinputs). The two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms ---  is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan1988].

@@

![Demonstration of how two inputs of different frequencies are processed by a nonlinear transducer. Additional harmonics are apparent in the Fourier spectrum (middle panel), and the contrast response function is suppressed (right panel).](){#fig-twoinputs}
:::
::: {.content-visible unless-format="docx"}
![Demonstration of how two inputs of different frequencies are processed by a nonlinear transducer. Additional harmonics are apparent in the Fourier spectrum (middle panel), and the contrast response function is suppressed (right panel).](Figures/4_twoinputs.pdf){#fig-twoinputs}
:::

@@

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells [@Kuffler1957; @Alpern1971], and cells with similar properties in the LGN [@Levitt2001] and cortex [@Movshon1975; @Hubel1959], will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position [@Hubel1962]. Both the amplitude and the phase of the SSVEP response to a single contrast-reversing sine-wave grating therefore contains information about nonlinear computations performed across a range of retinal and cortical cell types.

@@

Similar changes to the contrast response function might also be obtained using adaptation paradigms, in which the visual system is exposed to high contrast stimuli for long durations. Psychophysically, adaptation increases detection thresholds, but has little effect on contrast discrimination performance [@Ross1993], much like pattern masks [@Foley1994]. Although SSVEP adaptation effects show strong tuning for orientation [@Vergeer2018; @Campbell1970] and spatial frequency [@Mecacci1976], there appear to be very measurements of the full SSVEP contrast response function before and after an extended period of adaptation. One exception is a study by Bach et al [@BachGreenleeBuhler1988] that examined the effects of adaptation on SSVEP responses who found a paradoxical increase in the 1F SSVEP amplitude following prolonged exposure to high-contrast stimuli across much of the probe contrast range. They explain this as being due to the presence of a nonlinear transducer function that saturates at high probe contrast driving the response to higher harmonic frequencies. Adaptation reduces the apparent contrast of the input, leading to less distortion, a more sinusoidal output and a concentration of the frequency-tagged response at lower harmonics. This explanation is conceptually related to the 'supersaturation' effect that we observe in Figure 4.

@@

Because SSVEP responses at detection threshold are very small, estimating a threshold is achieved by measuring the contrast response function at relatively high levels, and extrapolating back along the function (either contrast vs response measured at a constant spatial frequency or spatial frequency vs response at a constant contrast level) to estimate its intercept with the x-axis (see @fig-sweepvep). This contrast level was shown to correspond approximately with psychophysically measured detection thresholds [@Norcia1986].

@@

![Sweep VEP simulation showing how a contrast detection threshold can be estimated from sweep VEP data measured at higher contrasts. The solid line is the regression fit to the lowest four data points, and the dashed line extrapolates the fit back to determine the contrast value when y=0, which gives a threshold estimate (red point).](){#fig-sweepvep}
:::
::: {.content-visible unless-format="docx"}
![Sweep VEP simulation showing how a contrast detection threshold can be estimated from sweep VEP data measured at higher contrasts. The solid line is the regression fit to the lowest four data points, and the dashed line extrapolates the fit back to determine the contrast value when y=0, which gives a threshold estimate (red point).](Figures/5_sweepplot.pdf){#fig-sweepvep}
:::

A robust estimate of the threshold therefore requires the measurement of the SSVEP amplitude at many different super-threshold contrast levels. This was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@Tyler1979]. To avoid hysteresis or short-term adaptation effects, the sweep is sometimes conducted both up and down in the same experiment [@Tyler1979; @Norcia1985a; @Norcia1990]. The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004; @Hoffmann2017; @Bach2008; @Bach2020].

This approach has revealed much about the development of visual abilities in infants [@Harris1976; @Braddick1986; @Atkinson1979]. In general, SSVEP measurements of infant vision have revealed that contrast sensitivity for both achromatic and chromatic contrast as well as stereoscopic depth perception develops earlier than had been supposed previously based on behavioural readouts [@Dobson1978; @Norcia1985b] with both chromatic and achromatic contrast detection reaching near-adult levels by around six months. Spatial acuity as measured by SSVEP reaches adult levels more slowly, but near-adult levels are recorded around one year [@Norcia1985b; @Hamer1989] compared to around six to seven years with behavioural measures [@Ellemberg1999; @Atkinson1983]. At least some of this difference is likely due to the relative objectivity and high SNR of the SSVEP technique compared to other methods such as preferential looking, which require careful measurement of the infant's gaze direction. However it should be noted that other groups have reported electrophysiological correlates of visual acuity that more closely match the behavioural measures [@De-Vries-Khoe1982].

The SSVEP technique has also been used to study the development of the contrast gain control mechanisms described in the previous section [@Candy2001; @Pei2017]. Although response- [@Morrone1986] or contrast gain control [@Candy2001; ; @Skoczenski1998] is measurable in infants as young as six weeks old, its development appears to be slower, with adult levels being reached at approximately 11 years [@Pei2017].

>>>>

To illustrate the effect of contrast gain control [@Heeger1992], we next include a second component (a 'mask') that contributes to the gain pool of the first (the 'target'). The mask stimulus will suppress the response at the target frequency, reducing its amplitude. The suppression is reciprocal - activity at the mask frequency is also reduced by the presence of the target, in a contrast-dependent manner [@Regan1988, @brownDevelopmentRivalryDichoptic1999, @Candy2001, @Busse2009]. At a single pair of (matched) contrast levels, we see a complex pattern of intermodulation terms in the transduced waveforms (left panel of @fig-twoinputs) and in the Fourier spectrum (middle panel of @fig-twoinputs). The two components interact generating nonlinear intermodulation terms at sums and differences of the input frequencies (F1, F2). The nature of the interaction --- and therefore the pattern of intermodulation terms ---  is determined by the computations happening as the contrast signal moves from retina to cortex [@Regan1988].

@@

![Demonstration of how two inputs of different frequencies are processed by a nonlinear transducer. Additional harmonics are apparent in the Fourier spectrum (middle panel), and the contrast response function is suppressed (right panel). The effects of using different nonlinearities can be explored in the accompanying code.](){#fig-twoinputs}
:::
::: {.content-visible unless-format="docx"}
![Demonstration of how two inputs of different frequencies are processed by a nonlinear transducer. Additional harmonics are apparent in the Fourier spectrum (middle panel), and the contrast response function is suppressed (right panel). The effects of using different nonlinearities can be explored in the accompanying code.](Figures/4_twoinputs.pdf){#fig-twoinputs}
:::

@@

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells [@Kuffler1957; @Alpern1971], and cells with similar properties in the LGN [@Levitt2001] and cortex [@Movshon1975; @Hubel1959], will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion [@McKeefry1996]. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position [@Hubel1962]. Both the amplitude and the phase of the SSVEP response to a single contrast-reversing sine-wave grating therefore contains information about nonlinear computations performed across a range of retinal and cortical cell types.

@@

Similar changes to the contrast response function might also be obtained using adaptation paradigms, in which the visual system is exposed to high contrast stimuli for long durations. Psychophysically, adaptation increases detection thresholds, but has little effect on contrast discrimination performance [@Ross1993], much like pattern masks [@Foley1994]. Although SSVEP adaptation effects show strong tuning for orientation [@Vergeer2018; @Campbell1970] and spatial frequency [@Mecacci1976], there appear to be very measurements of the full SSVEP contrast response function before and after an extended period of adaptation. One exception is a study by Bach et al [@bachGreenleeBuhler1988] that examined the effects of adaptation on SSVEP responses who found a paradoxical increase in the 1F SSVEP amplitude following prolonged exposure to high-contrast stimuli across much of the probe contrast range. They explain this as being due to the presence of a nonlinear transducer function that saturates at high probe contrast driving the response to higher harmonic frequencies. Adaptation reduces the apparent contrast of the input, leading to less distortion, a more sinusoidal output and a concentration of the frequency-tagged response at lower harmonics. This explanation is conceptually related to the 'supersaturation' effect that we observe in Figure 4.

@@

Because SSVEP responses at detection threshold are very small, estimating a threshold is achieved by measuring the contrast response function at relatively high levels (in the linear part of the log-contrast response function prior to saturation), and extrapolating back along the function (either contrast vs response measured at a constant spatial frequency or spatial frequency vs response at a constant contrast level) to estimate its intercept with the x-axis [@Campbell1970] (see @fig-sweepvep). This contrast level has been shown to correspond approximately with psychophysically measured detection thresholds [@campbellVisualEvokedPotential1972, @Norcia1986]. One factor affecting the accuracy of this technique is the presence of noise in the EEG signal which will increase the apparent amplitude of the SSVEP responses across the measurement range. The effect of this noise is particulary apparent for recordings with low SNR or where the slope of the CRF is shallow (and therefore where small changes in the vertical offset of the fit result in large changes in the X-axis intercept). Accurate estimates of thresholds therefore require the experimenter to model this noise and compensate for its effects - as detailed in [@Norcia1989].

@@

![Sweep VEP simulation showing how a contrast detection threshold can be estimated from sweep VEP data measured at higher contrasts. The solid line is the linear regression fit to the lowest four data points prior to the satuating part of the curve, and the dashed line extrapolates the fit back to determine the contrast value when y=0, which gives a threshold estimate (red point).](){#fig-sweepvep}
:::
::: {.content-visible unless-format="docx"}
![Sweep VEP simulation showing how a contrast detection threshold can be estimated from sweep VEP data measured at higher contrasts. The solid line is the linear regression fit to the  lowest four data points prior to the satuating part of the curve, and the dashed line extrapolates the fit back to determine the contrast value when y=0, which gives a threshold estimate (red point).](Figures/5_sweepplot.pdf){#fig-sweepvep}
:::

A robust estimate of the threshold therefore requires the measurement of the SSVEP amplitude at many different super-threshold contrast levels. This was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@reganRapidObjectiveRefraction1973,@Tyler1979]. To avoid hysteresis or short-term adaptation effects, the sweep is sometimes conducted both up and down in the same experiment [@Tyler1979; @Norcia1985a; @Norcia1990]. The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004; @Hoffmann2017; @Bach2008; @Bach2020].

The approach of estimating visual thresholds by extrapolating SSVEP responses at higher stimulus levels has revealed much about the development of visual abilities in infants [@Harris1976; @Braddick1986; @Atkinson1979]. In general, SSVEP measurements of infant vision have revealed that contrast sensitivity for both achromatic and chromatic contrast as well as stereoscopic depth perception develops earlier than had been supposed previously based on behavioural readouts [@Dobson1978; @Norcia1985b] with both chromatic and achromatic contrast detection reaching near-adult levels by around six months. Spatial acuity as measured by SSVEP reaches adult levels more slowly, but near-adult levels are recorded around one year [@Norcia1985b; @Hamer1989] compared to around six to seven years with behavioural measures [@Ellemberg1999; @Atkinson1983]. At least some of this difference is likely due to the relative objectivity and high SNR of the SSVEP technique compared to other methods such as preferential looking, which require careful measurement of the infant's gaze direction. However it should be noted that other groups have reported electrophysiological correlates of visual acuity that more closely match the behavioural measures [@De-Vries-Khoe1982].

The SSVEP technique has also been used to study the development of the contrast gain control mechanisms described in the previous section. Although response- [@Morrone1986] or contrast gain control [@Candy2001; ; @Skoczenski1998] is measurable in infants as young as six weeks old, its development appears to be slower, with adult levels being reached at approximately 11 years [@Pei2017].
